The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-01-26 article/article.tex
rework always-go-left

We will discuss two different multiple-choice placement strategies: The \emph{Greedy} algorithm (Section \ref{sec:greedy}) belongs to class 1, whereas the \emph{Always-Go-Left} scheme (Section \ref{sec:AlwaysGoLeft}) uses a non-uniform and independent selection of $d$ bins and thus belongs to class 2.

@@

Consider a variation of the above \emph{one-choice balls-into-bins} strategy, which we will call the \emph{Greedy scheme}: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}. If there are multiple bins with the same minimal load, chose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

The \emph{Greedy scheme} achieves a maximum load of 
\[
\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d} \right)
\]
for all $m$ w.h.p. \cite{ABKU99} \cite{BCSV06}.

@@

\subsubsection{History}
\label{sec:historyOfGreedy}
The history of proving this remarkable property of the two-choice paradigm is in itself quite interesting. It was first proven by Azar et al. \cite{ABKU99} for the case $m = n$. In the same paper, they claimed that the bound holds for arbitrary $m$. Unfortunately, their analysis breaks down for the heavily loaded case, where $m = \omega\left(n\right)$. Berenbrink at. all \cite{BFZR08} were the first to proof that indeed the same bound holds for the heavily loaded case\footnote{However, they use a slightly different definition for w.h.p.: An event $\mathcal E$ occurs with high probability if $\Pr\left[\mathcal E \right]  \geq 1- n^{-\alpha}$, for an arbitrarily chosen constant $\alpha \geq 1$.}. A simplified proof may be found in \cite{TW13}. Curiously, the case $d = 2$ was implicitly proven by Karp et al. in the context of simulating PRAMs on distributed memory machines\cite{KLM92}. As we will see in the next section, V\"ocking was able to slightly improve the bound as well as to show that the always-go-left strategy that he introduces is optimal up to additive constants \cite{VOC03}.

\subsection{The Always-Go-Left scheme}
\label{sec:AlwaysGoLeft}
The multiple-choice balls-into-bins strategy presented in Section \ref{sec:multiple-choice} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking \cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice balls-into-bins strategy with a \emph{nonuniform and independent} selection of the $d$ bins, such as the \emph{Always-Go-Left} algorithm introduced in \cite{VOC03}, achieves better\cite{BCSV06} load balancing than the \emph{Greedy} scheme (Section \ref{sec:greedy}). Furthermore, V\"ocking proved that the \emph{Always-Go-Left} algorithm is almost optimal \todo{be more precise}.

\todo{algorithm}

>>>>

We will discuss two different multiple-choice placement strategies: The \emph{Greedy} algorithm (Section \ref{sec:greedy}) belongs to class 1, whereas the \emph{Always-go-left} scheme (Section \ref{sec:AlwaysGoLeft}) uses a non-uniform and independent selection of $d$ bins and thus belongs to class 2.

@@

Consider a variation of the above \emph{one-choice balls-into-bins} strategy, which we will call the \emph{Greedy} scheme: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}. If there are multiple bins with the same minimal load, chose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

The \emph{Greedy} scheme achieves a maximum load of 
\[
\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d} \right)
\]
for all $m$ w.h.p. \cite{ABKU99} \cite{BCSV06}.

@@

\subsubsection{History}
\label{sec:historyOfGreedy}
The history of proving this remarkable property of the two-choice paradigm is in itself quite interesting. It was first proven by Azar et al. \cite{ABKU99} for the case $m = n$. In the same paper, they claimed that the bound holds for arbitrary $m$. Unfortunately, their analysis breaks down for the heavily loaded case, where $m = \omega\left(n\right)$. Berenbrink at. all \cite{BFZR08} were the first to proof that indeed the same bound holds for the heavily loaded case\footnote{However, they use a slightly different definition for w.h.p.: An event $\mathcal E$ occurs with high probability if $\Pr\left[\mathcal E \right]  \geq 1- n^{-\alpha}$, for an arbitrarily chosen constant $\alpha \geq 1$.}. A simplified proof may be found in \cite{TW13}. Curiously, the case $d = 2$ was implicitly proven by Karp et al. in the context of simulating PRAMs on distributed memory machines\cite{KLM92}. As we will see in the next section, V\"ocking was able to slightly improve the bound as well as to show that the \emph{Always-go-left} strategy that he introduces is optimal up to additive constants \cite{VOC03}.

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The \emph{Greedy} strategy presented in Section \ref{sec:multiple-choice} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking \cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice balls-into-bins strategy with a \emph{non-uniform} and independent selection of the $d$ bins, such as the \emph{Always-go-left} algorithm introduced in \cite{VOC03}, achieves better\cite{BCSV06} load balancing than the \emph{Greedy} scheme (Section \ref{sec:greedy}). Furthermore, V\"ocking proved that the \emph{Always-go-left} algorithm is optimal up to additive constants.

The \emph{Always-go-left} algorithm works as follows. Partition the $n$ bins into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\lceil\frac{n}{d}\rceil$ . We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, put the ball into the bin with smallest index $i$.
